Information Distribution Depends on Language-Specific Features

### 2019-01-29 cogsci/position-info.Rmd
new cogsci abstract & title

---
title: "Information Distribution Depends on a Language's Grammar and Sounds"
bibliography: library.bib
csl: apa6.csl
document-params: "10pt, letterpaper"

@@

abstract: >
    We provide evidence against the popular Uniform Information Density hypothesis (Levy & Jaeger, 2007) which proposes that information is transmitted at a constant rate close to channel capacity in human communication. Using a method based on the original Genzel & Charniak (2002) entropy measure, we construct a word-level model for entropy applicable to both spoken and written corpora. We apply this model to corpora from Wikipedia in well over a hundred languages. We find that the Uniform Information Density hypothesis does not predict the outcome of our word-level entropy model, but also that the by-word entropy distribution of a language is related to the typological features of the language. We use this evidence to suggest that people do not communicate information at a uniform rate, but that information distribution varies from language to language based on phonogical, morphological and syntactic features.

@@

We also ran this analysis on Spanish and Mandarin corpora from CHILDES. We used the Shiro corpus for Spanish (Shiro, 2000), which contains prompted narratives individually collected from over a hundred Venezualan schoolchildren, half from high SES backgrounds and half from low SES backgrounds. We used the Zhou dinner corpus for Mandarin Chinese (Li & Zhou, 2015), which contains dinner conversations between 5 to 6-year-old children and their parents collected in Shanghai.

For each corpus, we accessed transcripts of the corpus provided through the TalkBank system and computed over Roman alphabet transcriptions or transliterations of the original transcriptions. For Mandarin, we used pinyin transliterations of the utterances in the corpus with demarcated word boundaries, and for Japanese we used romanji (Roman alphabet) transliterations of words in the corpus. The Chinese characters used for writing Mandarin do not normally demarcate word boundaries by spacing words apart, and for normal Chinese writing including spaces between word boundaries can have a negative effect on reading times (Bai et al, 2008).

@@

Using Giuseppe Attardi's Wikiextractor tool \footnote{https://github.com/attardi/wikiextractor}, we extracted the text corpora for $165$ languages from Wikipedia by downloading a stored collection of Wikipedia entries in each langauge and randomly selecting several thousand articles from each Wikipedia language. Each language corpus was cleaned and limited to sentences between $6$ and $50$ words. Similar to our process for the Switchboard and CHILDES spoken corpora, we divided each corpus by sentence length, and then computed the unigram entropy measure on each word position within each sentence length. We wanted to directly compare and classify the unigram entropy curves of the languages from Wikipedia. We computed three slope treatments of each curve. In the *absolute* treatment, with sentence length denoted as $n$, we computed the slope between positions $1$ and $2$, positions $2$ and $3$, positions $3$ and $n-2$, positions $n-2$ and $n-1$ and positions $n-1$ and $n$. For the short utterances appearing the CHILDES speech corpora we examined, these appeared to be important junctions in the distributions, with a seeming plateau in the middle of the unigram entropy curve for each of the language corpora we examined in CHILDES.

@@

In the *absolute* and *relative 5* treatments, each language is embedded in $5$-dimensional space. In the *relative 10* treatment, each language is embedded in $10$-dimensional space. This allows for direct comparison between languages on Wikipedia using cosine similarity and moreover for unsupervised clustering analysis to compare the results of the Wikipedia entropy analysis to known typological features within the langauges in the Wikipedia dataset. To determine which phonological, morphological and syntactic features affected the embedding of a language in the Wikipedia dataset, we looked at the list of $144$ linguistic features in World Atlas of Language Structures (Dryer & Haspelmath, 2013). We limited the languages in the WALS database to only those in our Wikipedia dataset and performed missing-value imputation to obtain the features not coded in WALS for the languages from Wikipedia.

One approach is to cluster the languages used in the Wikipedia analysis on the basis of WALS features and then directly compare the WALS clustering results with the results of the hierarchical clustering on the Wikipedia slope data for each of the different treatments using clustering similarity evaluation methods such as the Rand index (Rand, 1971). However, the problem then arises of which combination of WALS features and how many features to include in the clustering analysis. This is a computationally intractable operation. Additionallly, deciding how many clusters to use in an unsupervised clustering analysis is an unsolved problem in machine learning.

@@

Cosine distance played some role in the feature determination. The top features were $83$, $81$, $95$ and $97$, which all characterize word order. This makes intuitive sense, and indicates that for this small subset of features and subset of languages that the embeddings in the slope space are related to the WALS features. Therefore typological features play a role in determining the positional entropy values for a language. This means that the entropy curves are in some way structured by the syntactic, morphological and phonological features of a language.

# Discussion 
In this paper, we have characterized and applied a model based on Genzel & Charniak (2002) for distinguishing entropy at each word position within sentences. In doing so, we have provided evidence against the Uniform Information Density hypothesis. Our work indicates that information is not distributed uniformly throughout an utterance regardless of languages, but that languages possess a characteristic information distribution determined by their syntactic, morphological and phonological features.

Our work has focused on words within sentences in both speech and text across typologically diverse languages. Other researchers have explored information rate cross-linguistically as a single number characterizing each language, most notably including Aylett & Turk (2004). That work characterizes information rate as information per unit time, and is complementary to the work we accomplish here. Languages may possess both an overall information transfer rate, showing the average amount of semantic information transmitted per syllable or word, and an entropy distribution indicative of the language's information distribution.

Research has indicated the effects of surprisal on fixation duration during eye-tracking studies (Demberg & Keller, 2008) (Boston et al., 2008) (Smith & Levy, 2013). Higher surprisal of a word is on average predictive of higher fixation duration. The *wrap-up effect* states that when a person reads written text, he or she will process sentence-final words more slowly on average then sentence-medial or sentence-initial words, due to integrating information from the entire sentence to form a final understanding of the sentence's meaning (Stowe et al., 2018) (Kuperman et al., 2010). The wrap-up effect is drawn from evidence in English, German and Dutch, all of which are languages with a large final increase in their entropy curve from our study. We aim to see in the future if as a consequence of our work in this paper, the wrap-up effect falls out from the final jump in the entropy curves in some languages. We would analyze eye-tracking data from languages with low positional entropy in sentence-final positions. Since the lexical surprisal metric in eye-tracking and our positional entropy metric are both measures of log-probability surprisal, then in languages such as Hindi where the sentence-final positional entropy values are low we expect to find no wrap-up effect.

>>>>

---
title: "Information Distribution Depends on Language-Specific Features"
bibliography: library.bib
csl: apa6.csl
document-params: "10pt, letterpaper"

@@

abstract: >
    We argue for a reformulation of the Uniform Information Density hypothesis @levy2007 in which the grammatical, phonological and morphological features of a language determine how information is distributed throughout utterances. We expand on a method from @yu2016 to compute entropy based on word position within utterances as a derivation of information distribution. We analyze adult and child spoken corpora as well as text corpora drawn from a variety of languages on Wikipedia. We find that each language has a robust, characteristic entropy curve across word position, and present evidence that these entropy curves are in part determined by typological features. We conclude that speakers may tend towards communicating information uniformly, but the distribution of information in their utterances is conditioned on the language they communicate in.

@@

We also ran this analysis on Spanish and Mandarin corpora from CHILDES. We used the Shiro corpus for Spanish @shiro2000, which contains prompted narratives individually collected from over a hundred Venezualan schoolchildren, half from high SES backgrounds and half from low SES backgrounds. We used the Zhou dinner corpus for Mandarin Chinese (Li & Zhou, 2015), which contains dinner conversations between 5 to 6-year-old children and their parents collected in Shanghai.

For each corpus, we accessed transcripts of the corpus provided through the TalkBank system and computed over the Latin alphabet transcriptions or transliterations of the original transcriptions. For Mandarin, we used pinyin transliterations of the utterances in the corpus with demarcated word boundaries. The Chinese characters used for writing Mandarin do not normally demarcate word boundaries by spacing words apart, and for normal Chinese writing including spaces between word boundaries can have a negative effect on reading times @bai2008.

@@

Using Giuseppe Attardi's Wikiextractor tool \footnote{https://github.com/attardi/wikiextractor}, we extract text corpora from Wikipedia by downloading a stored collection of Wikipedia entries in each langauge and randomly selecting several thousand articles from each Wikipedia language. Each language corpus was cleaned and limited to sentences between $6$ and $50$ words. Similar to our process for the Switchboard and CHILDES spoken corpora, we divided each corpus by sentence length, and then computed the unigram entropy measure on each word position within each sentence length. We wanted to directly compare and classify the unigram entropy curves of the languages from Wikipedia. We computed three slope treatments of each curve. In the *absolute* treatment, with sentence length denoted as $n$, we computed the slope between positions $1$ and $2$, positions $2$ and $3$, positions $3$ and $n-2$, positions $n-2$ and $n-1$ and positions $n-1$ and $n$. For the short utterances appearing the CHILDES speech corpora we examined, these appeared to be important junctions in the distributions, with a seeming plateau in the middle of the unigram entropy curve for each of the language corpora we examined in CHILDES.

@@

In the *absolute* and *relative 5* treatments, each language is embedded in $5$-dimensional space. In the *relative 10* treatment, each language is embedded in $10$-dimensional space. This allows for direct comparison between languages on Wikipedia using cosine similarity and moreover for unsupervised clustering analysis to compare the results of our position-wise analysis on datasets from Wikipedia to known typological features within the languages. To determine which phonological, morphological and syntactic features affected the embedding of a language in the Wikipedia dataset, we use the linguistic features in World Atlas of Language Structures (Dryer & Haspelmath, 2013).

[Ideally do everything, but missing values in WALS, and how to compare every combination of features] An ideal approach using all of the corpora we can download from Wikipedia and all $144$ features from the WALS database. Cluster the languages used in the Wikipedia analysis on the basis of WALS features and then directly compare the WALS clustering results with the results of the hierarchical clustering on the Wikipedia slope data for each of the different treatments using clustering similarity evaluation methods such as the Rand index (Rand, 1971). However, the problem then arises of which combination of WALS features and how many features to include in the clustering analysis. This is a computationally intractable operation. Additionallly, deciding how many clusters to use in an unsupervised clustering analysis is an unsolved problem in machine learning.

@@

Cosine distance played some role in the feature determination. The top four features all characterize word order, which indicates that for this subset of features and languages that the embeddings in the slope space are related to the WALS features. Therefore typological features play a role in determining the positional entropy values for a language. This means that the entropy curves are in some way structured by the syntactic, morphological and phonological features of a language.

# Discussion 
In this paper, we have extended and applied a model from @yu2016 and derived from @genzel2002 for distinguishing entropy at each word position within sentences. We have argued that the outcomes of this model are derived from the information distribution of a language. As this model varies cross-linguistically in terms of

Follow-up with WALS features using missing-value imputation. However here are some of the problems with that.

Our work complements the approach of studies such as @aylett2004, where languages are characterized by a rate of semantic information per syllable. This body of work attests to languages possessing different rates of information transfer based on typological features as well.

We also propose follow-up within linguistic processing using evidence from cross-linguistic eye-tracking. Research has indicated the effects of surprisal on fixation duration during eye-tracking studies [@demberg2008; @boston2008; @smith2013]; higher surprisal of a word is on average predictive of higher fixation duration. The *wrap-up effect* states that when a person reads written text, he or she will process sentence-final words more slowly on average then sentence-medial or sentence-initial words, due to integrating information from the entire sentence to form a final understanding of the sentence's meaning [@stowe2018; @kuperman2010]. The wrap-up effect is drawn from evidence in languages with a large final increase in their entropy curve from our study, so perhaps the wrap-up effect derives from the sentence-final increase in entropy curves observed in English and other Germanic languages.
